*Prochlorococcus marinus* responses to light and oxygen

### 2024-02-23 Code/Manuscript_O2.Rmd
Fixed citations

Low-Light clades thrive in deeper ocean waters, extending beyond 200 m in depth [@partenskyProchlorococcusMarinePhotosynthetic1999], where only ~1% of the surface irradiance penetrates, primarily in the blue (450 nm) to green (520 nm) spectral range [@holtropVibrationalModesWater2021]. Clade LLIV, including cultured strain MIT9313, falls near the base of the *Prochlorococcus* radiation, and is characterized by preference for low light, typical of depths from 120 m to 200 m (Partensky et al., 1999). LLIV members are, as yet, the only cultured strains to have been found in Oxygen Minimum Zones (OMZ).  Some as yet uncultured *P. marinus* strains in clades LLV and LLVI also thrive in OMZ of the subtropical Atlantic and Pacific Oceans, where dissolved oxygen concentrations [O~2~] can be less than 20 µM [@ulloaCyanobacteriumProchlorococcusHas2021; @lavinNovelLineagesProchlorococcus2010; @goerickeNovelNicheProchlorococcus2000; @robledoCrypticOxygenCycling2017] (Johnson et al., 1999). *P. marinus* LL ecotypes may indeed dominate the phytoplankton within OMZ
[@lavinNovelLineagesProchlorococcus2010;
@ulloaCyanobacteriumProchlorococcusHas2021; @partenskyProchlorococcusAdvantagesLimits2010], and may be net O~2~ consumers (2018, Partensky et al. 2018).

@@

Current niches of *P. marinus* span across vertical ocean water columns
[@chisholmNovelFreelivingProchlorophyte1988; @veldhuisVerticalDistributionPigment1990; @partenskyProchlorococcusMarinePhotosynthetic1999] and extend into regions with varying dissolved oxygen concentrations
[@partenskyProchlorococcusAdvantagesLimits2010; @goerickeNovelNicheProchlorococcus2000] and even into anoxic oceans [@lavinNovelLineagesProchlorococcus2010], [@ulloaCyanobacteriumProchlorococcusHas2021; @garcia-robledoCrypticOxygenCycling2017a]. Climate change is drastically altering the dissolved oxygen concentrations ([O~2~]) of our ocean waters. Research indicates that the oxygen minimum zones (OMZs) in the Pacific and Indian Oceans are expanding [@garcia-sotoOverviewOceanClimate2021; @buseckeDivergingFatesPacific2022] and that the levels of dissolved oxygen (O~2~) in our global oceans are generally decreasing [@matearLongtermChangesDissolved2003; @garcia-sotoOverviewOceanClimate2021], with significant decreases observed toward the poles [@helmObservedDecreasesOxygen2011].

>>>>

Low-Light clades thrive in deeper ocean waters, extending beyond 200 m in depth [@partenskyProchlorococcusMarinePhotosynthetic1999], where only ~1% of the surface irradiance penetrates, primarily in the blue (450 nm) to green (520 nm) spectral range [@holtropVibrationalModesWater2021]. Clade LLIV, including cultured strain MIT9313, falls near the base of the *Prochlorococcus* radiation, and is characterized by preference for low light, typical of depths from 120 m to 200 m (Partensky et al., 1999). LLIV members are, as yet, the only cultured strains to have been found in Oxygen Minimum Zones (OMZ).  Some as yet uncultured *P. marinus* strains in clades LLV and LLVI also thrive in OMZ of the subtropical Atlantic and Pacific Oceans, where dissolved oxygen concentrations [O~2~] can be less than 20 µM [@ulloaCyanobacteriumProchlorococcusHas2021; @lavinNovelLineagesProchlorococcus2010; @goerickeNovelNicheProchlorococcus2000; @garcia-robledoCrypticOxygenCycling2017] (Johnson et al., 1999). *P. marinus* LL ecotypes may indeed dominate the phytoplankton within OMZ
[@lavinNovelLineagesProchlorococcus2010;
@ulloaCyanobacteriumProchlorococcusHas2021; @partenskyProchlorococcusAdvantagesLimits2010], and may be net O~2~ consumers (2018, Partensky et al. 2018).

@@

Current niches of *P. marinus* span across vertical ocean water columns
[@chisholmNovelFreelivingProchlorophyte1988; @veldhuisVerticalDistributionPigment1990; @partenskyProchlorococcusMarinePhotosynthetic1999] and extend into regions with varying dissolved oxygen concentrations
[@partenskyProchlorococcusAdvantagesLimits2010; @goerickeNovelNicheProchlorococcus2000] and even into anoxic oceans [@lavinNovelLineagesProchlorococcus2010], [@ulloaCyanobacteriumProchlorococcusHas2021; @garcia-robledoCrypticOxygenCycling2017]. Climate change is drastically altering the dissolved oxygen concentrations ([O~2~]) of our ocean waters. Research indicates that the oxygen minimum zones (OMZs) in the Pacific and Indian Oceans are expanding [@garcia-sotoOverviewOceanClimate2021; @buseckeDivergingFatesPacific2022] and that the levels of dissolved oxygen (O~2~) in our global oceans are generally decreasing [@matearLongtermChangesDissolved2003; @garcia-sotoOverviewOceanClimate2021], with significant decreases observed toward the poles [@helmObservedDecreasesOxygen2011].
